From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-13 paper/paper.Rmd
mostly refs

csl: ecology-letters.csl
bibliography: refs.bib
output: rticles::elsevier_article  
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5)
library(tidyverse)
library(gridExtra)
library(ggthemes)
colours <- ptol_pal()(2)
```

@@

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: ecology is complex, and much effort
has been spent on illustrating the importance that complexity: age and stage
structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
While acknowledging more realistic examples and more general results where avialable,
using specific and familiar simple models that -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@DeValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

This does not attempt to summarize all of the literature on the origins and
types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskienen2010; @Black2012; @Schrieber2012; @Lenormand2009]

@@

What is noise? Where does it come from?  It is important to bear
in mind that whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
Simply put, *noise is all that we leave out of our model*.  Early
quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models [@Black2012; @Ovaskainen2010].

@@

@@

Understanding the origin of sustained oscillations has long been a
central question in ecology [@Hastings1996; @Bjornstad2001].  The
simple Lotka Volterra model produces only neutrally stable centers,
while the more realistic model that introduces carrying capacity
for prey as they compete for resources results in a model that
produces only damped oscillations, eventually settling into a stable
state (Figure \ref{quasicycles}A).  However, the introduction of relatively small
noise into the system can be sufficient to excite the system into
sustained oscillations (Figure \ref{quasicycles}B).  In the example shown here, a
simple predator-prey system is driven by environmental white noise
$\xi_t$, with prey following the equation:

@@

Early work by @Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod, while @Black2010 illustrates the use of
power-spectrum approaches to examine the behavior of quasi-cycles
in common disease outbreak models, focusing on a case where noise
arises explicitly from demographic stochasticity, rather than generic
environmental noise as in the model above and elsewhere.  @Pineda-Krch2007
compare patterns from these stochastically generated cycles to
stable limit cycles they resemble.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

@@

The challenge of identifying warning signs that a system is approaching
a tipping point has proven a rich area for applications of stochastic
processes.  We have already seen how in addition to the patterns of
increasing variance and auto-correlation, our survey of stochastic phenomena
has highlighted the increased potential to observe stochastic switching,
or flickering, as the system approaches the tipping point.  Challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
The greater our suite of stochastic indicators, the better our ability
to pin down the processes involved.  Such examples suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussman2000; @Strogatz2001].

>>>>

## Use trends style to quickly count refs
#csl: trendsjournals.csl
csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = TRUE)
library(tidyverse)
library(gridExtra)
library(ggthemes)
colours <- ptol_pal()(2)
```

@@

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: ecology is complex, and much effort
has been spent on illustrating the importance that complexity: age and stage
structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
While acknowledging more realistic examples and more general results where avialable,
using specific and familiar simple models that -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

@@

What is noise? Where does it come from?  Simply put, *noise is all that we leave out of our model*. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochastic summary and underlying process more precise:
that is, on understanding the origins of noise [@Coulson2004].

Over the past several decades, research into stochastic modeling
has

This does not attempt to summarize all of the literature on the origins and
types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskainen2010; @Black2012; @Schreiber2012; @Lenormand2009]

justify simple models, 
citations adapting this to complex models:

Early
quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models [@Black2012; @Ovaskainen2010].

@@

Few topics have recieved greater attention then the question of
the persistence of a given species [@Kendall1949;  @Bartlett1959; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]

and coexistence of multiple species

[@Caswell1978; @Tuljapurkar1980; @Chesson1981; @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]

Structured populations 
[@Cohen1979; @Tuljapurkar1989; @Metcalf2015; @Vindenes2017b]

See in particular reviews by @Ovaskainen2010 and @Schreiber2012

Individual heterogeneity 
@Vindenes2015; @Hart2016

@Lee2017

@@

<!--  -->

Understanding the origin of sustained oscillations has long been a
central question in ecology [@Hastings1996; @Bjornstad2001].  The
simple Lotka Volterra model produces only neutrally stable centers,
while the more realistic model that introduces carrying capacity
for prey as they compete for resources results in a model that
produces only damped oscillations, eventually settling into a stable
state (Figure \ref{quasicycles}A).  However, the introduction of relatively small
noise into the system can be sufficient to excite the system into
sustained oscillations (Figure \ref{quasicycles}B) 
[e.g. empirical examples in @Blarer1999; @Greenman2003; @Black2010].  
In the example shown here, a
simple predator-prey system is driven by environmental white noise
$\xi_t$, with prey following the equation:

@@

Early work by @Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod, while @Black2010 illustrates the use of
power-spectrum approaches to examine the behavior of quasi-cycles
in common disease outbreak models, focusing on a case where noise
arises explicitly from demographic stochasticity, rather than generic
environmental noise as in the model above and elsewhere [@Kendall1999; @Pascual2003].  
@Pineda-Krch2007 compare patterns from these stochastically generated cycles to
stable limit cycles they resemble.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

@@

The challenge of identifying warning signs that a system is approaching
a tipping point has proven a rich area for applications of stochastic
processes.  We have already seen how in addition to the patterns of
increasing variance and auto-correlation, our survey of stochastic phenomena
has highlighted the increased potential to observe stochastic switching,
or flickering, as the system approaches the tipping point.  Challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
The greater our suite of stochastic indicators, the better our ability
to pin down the processes involved.  Such examples suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Strogatz2001].
